---
title: "A Simple Statistical Exploration of the Exponential Distribution"
author: "Wayne Witzke"
output: pdf_document
---

## Synopsis
This analysis explores the relationship between theoretical and sample
statistics for the exponential distribution. It simulates sets of experimental
samples drawn from the exponential distribution, and calculates statistics for
those sets. Finally, it shows that the distribution of sample means of those
sets is Gaussian, and thus the Central Limit Theorem applies.

## Preliminaries

This section prepares the analysis to run. For more information on replicating
this analysis, see the appendix.

```{r setup}
knitr::opts_chunk$set( echo = TRUE );
library(ggplot2);
set.seed(397413653);
```

## Simulation

This analysis generates 1000 sets of 40 random numbers taken from the
exponential distribution. The chosen rate ($\lambda$) is 0.2, giving the
distribution a theoretical mean and standard deviation of $\mu=\sigma=\frac{1}{\lambda}=5$.

```{r simulation}
sim.matrix = matrix( rexp( 40000, 0.2 ), 1000, 40 );
raw.mean = mean( as.vector( sim.matrix ) );
raw.sd = sd( as.vector( sim.matrix ) );
```

Note that, over all generated values, the mean is $\bar{x}=`r raw.mean`$ and
the standard deviation is $s=`r raw.sd`$.

## Results

### Sample Mean vs. Theoretical Mean
```{r sample_mean_vs_theoretical_mean}
sim.means = apply( sim.matrix, 1, mean );
sim.mean = mean( sim.means );
```
Once again, the theoretical mean of the distribution is given by 
$\mu=\frac{1}{\lambda}=5$. Taking the average of the means of each
simulation, we see that the simulated sample mean is $\bar{X}=`r sim.mean`$.

### Sample Variance vs. Theoretical Variance
```{r sample_variance_vs_theoretical_variance}
sim.variance = var( sim.means );
sim.sd = sd( sim.means );
```
The theoretical variance of the distribution is given by 
$\sigma^2=(\frac{1}{\lambda})^2=25$. The variance of the sample mean, however,
is a much smaller $s_m^2=`r sim.variance`$. Note that there was no expectation
that the sample mean variance and theoretical variance would be similar, as they
describe different distributions. However, note that $s_m^2\approx\frac{s^2}{n}=`r raw.sd^2/40`\approx\frac{\sigma^2}{n}=`r 25/40`$,
where $n=40$ is the sample size, $\frac{\sigma}{\sqrt{n}}$ is the true
standard error, and $\frac{s}{\sqrt{n}}$ is the sample standard error.

The statistics for the theoretical distribution and sample means are summarized
in the following table.
\begin{center}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{|l|c|c|}
 \hline
 \multicolumn{3}{|c|}{Theoretical Distribution vs. Simulated Sample Mean Statistics} \\
 \hline
 \hline
 & Mean & Variance \\
 \hline
 Theoretical Exponential Distribution & $\frac{1}{\lambda}=5$ & $\left ( \frac{1}{\lambda} \right ) ^2=25$ \\
 \hline
 Simulated Sample Means & `r sim.mean` & `r sim.variance` \\
 \hline
\end{tabular}
\egroup
\end{center}


### Distribution
As can be seen from the following histogram, random values drawn from the
exponential distribution do not have a normal distribution. For example, the
mean does not occur at the peak value.

```{r simulation_histogram, fig.width = 7, fig.height = 4}
raw.histogram =
    ggplot( show.legend = TRUE ) + aes( as.vector( sim.matrix ) ) +
    geom_histogram( binwidth = 0.8, col = "black", aes( fill = ..count.. ) ) +
    xlab( "Simulated Values" ) + ylab("Count") +
    ggtitle( "Distribution of Random Exponential Numbers" ) +
    geom_vline( aes( xintercept = raw.mean, color = "Mean" ), size = 1 ) +
    scale_color_manual( name = "", values = c( Mean = "red" ) ) +
    scale_fill_continuous( name = "Count" );
print( raw.histogram );
```

However, the distribution of the sample means calculated from the individual
simulations *is* normal. The following histogram shows the distribution of the
sample means and fits a normal curve with the same statistics, clearly showing
that the sample means follows a Gaussian distribution.

```{r sample_means_histogram, fig.width = 7, fig.height = 4}
sim.histogram =
    ggplot( show.legend = TRUE ) + aes( sim.means ) +
    geom_histogram( binwidth = 0.15, col = "black", aes( fill = ..count.. ) ) +
    xlab( "Sample Means" ) + ylab("Count") +
    ggtitle(
        "Distribution of Sample Means for Random Exponential Numbers"
    ) +
    geom_vline( aes( xintercept = sim.mean, color = "Mean" ), size = 1 ) +
    scale_color_manual( name = "", values = c( Mean = "red" ) ) +
    scale_fill_continuous( name = "Count" ) +
    stat_function(
        fun = function(x, mean, sd, n, bw) dnorm(x=x,mean=mean,sd=sd)*n*bw,
        color = "green3",
        size = 1.5,
        args = list( mean = sim.mean, sd = sim.sd, n = 1000, bw = 0.15 )
    );
print( sim.histogram );
```
The Central Limit Theorem states that

> The distribution of averages of independent and identically distributed
> variables becomes that of a standard normal as the sample size increases.

In this case, despite the underlying shape of the exponential distribution, the
distribution of the sample means clearly approaches a normal distribution,
centered about the mean of the exponential distribution, as predicted by the
Central Limit Theorem. Note that the variance on the sample mean is controlled
by the sample size, $n=40$.


\newpage
## Appendix

### System Information

This analysis was performed using the hardware and software specified in this
section.

```{r system_info}
sessionInfo();
```
