---
title: "An Analysis of the Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Wayne Witzke"
header-includes:
  - \usepackage[table]{xcolor}
output: pdf_document
---

## Synopsis
Our analysis examines whether dose (0.5 mg/day, 1 mg/day, 2 mg/day) or delivery
method (orange juice vs. ascorbic acid) has an effect on the tooth growth in
guinea pigs. We examine 6 individual groups of guinea pigs, each with a
different combination of dosage and delivery method, and determine if there are
any significant differences between these groups. We find that there is fairly
strong evidence that increasing dosage changes tooth growth, but that there is
less evidence that delivery method has a impact.

## Preliminaries

This section prepares the analysis to run. For more information on replicating
this analysis, see the appendix.

```{r setup}
library(ggplot2); library(plyr);
data(ToothGrowth);
options(digits=5);
```

## Exploratory Analysis

### Summary of Data

This analysis uses the `ToothGrowth` data set included natively with most R
distributions. As described in the R help document, each of 60 guinea pigs was
given a dose of 0.5, 1, or 2 mg/day of vitamin C by one of two delivery
methods, orange juice (coded `OJ`) or ascorbic acid (coded `VC`). The length of
their odontoblasts were then measured. The structure of the data set is:

```{r data_structure}
str(ToothGrowth)
```

### Exploratory Graphs

We are interested in seeing if there is a significant difference between the
doses, and between the delivery methods. Unfortunately, because of potential
correlations between dose and delivery method, we cannot simply compare
aggregates, so we will look at all 6 groups individually. Below are histograms
showing the distribution of all 60 samples in each of the 6 groups
individually, each with a Gaussian curve fit to show how well the sample
approximates a normal distribution.

```{r explore_graphs, fig.width=8, fig.height=3}
#   We want to fit normals to the graphs here.
ToothGrowthNorms =
    ddply( ToothGrowth, .(supp, dose), function(df) {
        lenSeq=seq(min(ToothGrowth$len)-8, max(ToothGrowth$len)+8, length=200 );
        data.frame(len=lenSeq, count=2.37*10*dnorm(lenSeq, mean(df$len), sd(df$len)));
    });
ToothGrowth.histogram =
    ggplot( ToothGrowth, aes( len ), show.legend = TRUE ) +
    geom_histogram( binwidth=2.37, col = "black", aes( fill = ..count.. ) ) +
    xlab( "Length" ) + ylab("Count") +
    ggtitle( "Exploratory Histograms for Guinea Pig Tooth Growth by Group" )  +
    facet_grid( supp ~ dose ) +
    scale_fill_continuous( name = "Count" ) +
    geom_line(aes(y = count), data=ToothGrowthNorms, color="green3", size=1);
print( ToothGrowth.histogram );
```

It is difficult to tell from these plots if every group is normal. However,
since most of them appear to be at least approximately normal, for this
analysis, we will assume that they follow the Student's t distribution.
This is appropriate both because of the small sample sizes ($n=10$) and because
we do not know the actual population standard deviation.

## Hypothesis Testing

We are interested in seeing if there is a difference between individual groups.
Our null hypothesis, $H_0$, is that there is no difference, and the alternative
hypothesis, $H_\alpha$, is that the dose or delivery method makes a difference.
Because of the potential for correlation betweens doses and delivery methods,
we will be looking for significant differences between any two groups, which is
$\eta=15$ hypothesis tests. To control for the number of tests performed, we will use
the *Bonferroni correction*. Our $\alpha=0.05$ for each test individually. With
the correction, however, we will be using an effective $\alpha_c=\frac{\alpha}{\eta}=`r 0.05/15`$ for
each hypothesis test. We will be using a two-sided Student's t test to generate
p-values, and we will not treat the variance between groups as equal.

The figure below summarizes the results of the hypothesis testing. Note that
p-values that reject the null hypothesis are highlighted in green. (See the
Appendix for the full code to generate this figure.)

```{r refactor, echo=FALSE}
TG = ToothGrowth;
TG$group = with(TG, interaction( dose, supp ) );
```

\begin{center}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{|r|c|c|c|c|c|}
 \hline
 \multicolumn{6}{|c|}{Summary of P-Values Calculated Across Groups} \\
 \hline
 \hline
 Dose/Mthd & 0.5/OJ & 1/OJ & 2/OJ & 0.5/VC & 1/VC \\
 \hline
 1/OJ & \cellcolor{green} $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="1.OJ","len"],alternative="two.sided")$p.value`$ & --- & --- & --- & --- \\
 \hline
 2/OJ & \cellcolor{green} $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="2.OJ","len"],alternative="two.sided")$p.value`$ & $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="2.OJ","len"],alternative="two.sided")$p.value`$ & --- & --- & --- \\
 \hline
 0.5/VC & $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="0.5.VC","len"],alternative="two.sided")$p.value`$ & \cellcolor{green} $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="0.5.VC","len"],alternative="two.sided")$p.value`$ & \cellcolor{green} $`r t.test(TG[TG$group=="2.OJ","len"],TG[TG$group=="0.5.VC","len"],alternative="two.sided")$p.value`$ & --- & --- \\
 \hline
 1/VC & $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="1.VC","len"],alternative="two.sided")$p.value`$ & \cellcolor{green} $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="1.VC","len"],alternative="two.sided")$p.value`$ & \cellcolor{green} $`r t.test(TG[TG$group=="2.OJ","len"],TG[TG$group=="1.VC","len"],alternative="two.sided")$p.value`$ & \cellcolor{green} $`r t.test(TG[TG$group=="0.5.VC","len"],TG[TG$group=="1.VC","len"],alternative="two.sided")$p.value`$ & --- \\
 \hline
 2/VC & \cellcolor{green} $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="2.VC","len"],alternative="two.sided")$p.value`$ & $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="2.VC","len"],alternative="two.sided")$p.value`$ & $`r t.test(TG[TG$group=="2.OJ","len"],TG[TG$group=="2.VC","len"],alternative="two.sided")$p.value`$ & \cellcolor{green} $`r t.test(TG[TG$group=="0.5.VC","len"],TG[TG$group=="2.VC","len"],alternative="two.sided")$p.value`$ & \cellcolor{green} $`r t.test(TG[TG$group=="1.VC","len"],TG[TG$group=="2.VC","len"],alternative="two.sided")$p.value`$ \\
 \hline
\end{tabular}
\egroup
\end{center}

## Conclusions

There are three sets of p-values that are of clear interest. The first is the
increase in dosage in orange juice, and the second is the increase in dosage
in ascorbic acid. Examining these p-values, there appears to be a significant
effect caused by increasing dosage. The only p-value in these two sets which
fails to reject $H_0$ is between 1/OJ and 2/OJ. In addition, though there may
be cross correlations that muddy the data, increase in dosage across delivery
methods also appears to generally support the hypothesis that differeing doses
has a significant impact on tooth growth.

The third set of p-values that interest us are the ones where dosage is held
constant, and only the delivery method changes. In this case, there appears to
be less evidence of a significant relationship. Only 1/OJ vs. 1/VC rejected
$H_0$, while 0.5/VC vs. 0.5/OJ and 2/VC vs 2/OJ did not. Still, we conclude
that there is likely some effect based on delivery method, though we would
probably urge additional testing to be certain.

These conclusions hold only if the underlying samples were normally distributed
within their own groups.

\newpage
## Appendix

### System Information

This analysis was performed using the hardware and software specified in this
section.

```{r system_info}
sessionInfo();
```

### Code to Generate the P-Value Summary Table
```{r appendix_table_code, eval=FALSE}

`r TG = ToothGrowth; TG$group = with(TG, interaction( dose, supp ) );`

\begin{center}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{|r|c|c|c|c|c|}
 \hline
 \multicolumn{6}{|c|}{Summary of P-Values Calculated Across Groups} \\
 \hline
 \hline
 Dose/Mthd & 0.5/OJ & 1/OJ & 2/OJ & 0.5/VC & 1/VC \\
 \hline
 1/OJ
 &  \cellcolor{green} $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="1.OJ","len"],
 alternative="two.sided")$p.value`$
 &  --- &  --- &  --- &  --- \\
 \hline
 2/OJ
 &  \cellcolor{green} $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="2.OJ","len"],
 alternative="two.sided")$p.value`$
 &  $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="2.OJ","len"],
 alternative="two.sided")$p.value`$
 &  --- &  --- &  --- \\
 \hline
 0.5/VC
 &  $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="0.5.VC","len"],
 alternative="two.sided")$p.value`$
 &  \cellcolor{green} $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="0.5.VC","len"],
 alternative="two.sided")$p.value`$
 &  \cellcolor{green} $`r t.test(TG[TG$group=="2.OJ","len"],TG[TG$group=="0.5.VC","len"],
 alternative="two.sided")$p.value`$
 &  --- &  --- \\
 \hline
 1/VC
 &  $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="1.VC","len"],
 alternative="two.sided")$p.value`$
 &  \cellcolor{green} $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="1.VC","len"],
 alternative="two.sided")$p.value`$
 &  \cellcolor{green} $`r t.test(TG[TG$group=="2.OJ","len"],TG[TG$group=="1.VC","len"],
 alternative="two.sided")$p.value`$
 &  \cellcolor{green} $`r t.test(TG[TG$group=="0.5.VC","len"],TG[TG$group=="1.VC","len"],
 alternative="two.sided")$p.value`$
 &  --- \\
 \hline
 2/VC
 &  \cellcolor{green} $`r t.test(TG[TG$group=="0.5.OJ","len"],TG[TG$group=="2.VC","len"],
 alternative="two.sided")$p.value`$
 &  $`r t.test(TG[TG$group=="1.OJ","len"],TG[TG$group=="2.VC","len"],
 alternative="two.sided")$p.value`$
 &  $`r t.test(TG[TG$group=="2.OJ","len"],TG[TG$group=="2.VC","len"],
 alternative="two.sided")$p.value`$
 &  \cellcolor{green} $`r t.test(TG[TG$group=="0.5.VC","len"],TG[TG$group=="2.VC","len"],
 alternative="two.sided")$p.value`$
 &  \cellcolor{green} $`r t.test(TG[TG$group=="1.VC","len"],TG[TG$group=="2.VC","len"],
 alternative="two.sided")$p.value`$ \\
 \hline
\end{tabular}
\egroup
\end{center}
```
